---
title: "Census and factors"
output: html_notebook
---

# Factors, and the UK census

Soon we will explore the data from the [2011 UK
census](https://www.ons.gov.uk/census/2011census) from the Office of [National
Statistics](https://www.ons.gov.uk) (ONS).

But to start, we explore R *factors*.

Factors are a special type of *vector* in R.

## About filling in this notebook

Remember to run the chunks of code as you go.

Some of the chunks are just for you to run.  Others ask you to fill in some
code.  These often have markers like `...` to tell you where to fill in the
code.

## Revision on vectors

As you recall, a vector is a sequence of values.  It can be a sequence of
numbers.  Here we use the `c` function to Combine the numbers into a vector.

```{r}
some_numbers <- c(5, 2, 3, 1, 4, 3, 2, -1, 1)
some_numbers
```

The vector has a `length`:

```{r}
length(some_numbers)
```

It can also be a sequence of strings.  Strings are bits of text.  To make
a string, we put the text we want between quotes, like:

```{r}
'this'
```

Here is a vector of strings.  Notice the quote characters around each string,
to tell R this is a bit of text, rather than a variable name or a function.

```{r}
some_strings <- c('good', 'bad', 'ugly', 'good', 'ugly')
some_strings
```

Make a new vector of strings called `disappointing_finale` that has the following strings:

* Star
* Wars
* The
* Rise
* of
* Skywalker

```{r}
disappointing_finale <- ...
# Show the number of words
length(disappointing_finale)
```

## Factors

Factors are special types of vectors.

Factors express categories.  They are important in several ways.  One way they
are important is telling R that the data identify categories.  R can then treat
these values differently in a variety of ways.

Imagine for example that I recorded my friends grades in A level geography with
numbers.  I decided to use the code 1 for an A, 2 for a B and so on.  -1 means
a U grade.  Yes, I have one friend who is not good at geography.

```{r}
friends_grades <- c(2, 4, 5, 1, 3, 3, 2, -1, 1)
# Show the vector of grades
friends_grades
```

Show the type of the `friends_grades` vector using `class`:

```{r}
...
```

I intend the numbers in `friends_grades` to express categories --- category
1 for people who got an A, 2 for people who got a B, and so on.

We can make a factor out of a vector by using the `factor` function.

```{r}
# Run this chunk
friends_factor <- factor(friends_grades)
friends_factor
```

Notice how the display changed, when R shows us the factor.  First R shows us
the numbers, as it did before, for `friends_grades`.  Below that, R shows the
"levels" of the factor.  Notice that the levels are also the unique values in
the vector.  The levels are the *categories* contained in the factor.  That is,
all the values of 1 in the vector are now in a category (level) "1", all the
values of 2 in the category (level) "2" and so on.  The levels are *labels* for
the categories. This might not be obvious yet.  We can see the levels of
a factor by using the `levels` function:

```{r}
# Run this chunk
levels(friends_factor)
```

Notice too that the levels are actually *strings* (bits of text).  The output
shows us the bit of text "-1" for the value -1, "1" for the value 1, and so on.

We often want to set more useful labels.

For example, we know that the values in `friends_factor` are codes for my
friends' grades at A-level: -1 for "U" grade, 1 for "A" and so on.

We can set the levels (labels) of a factor directly.  First we make a vector of
strings that have the labels we want:

```{r}
# A vector of strings with the labels we want
grades <- c('U', 'A', 'B', 'C', 'D', 'E', 'F')
grades
```

We use this vector to set the labels for the levels.

```{r}
# Set the labels for the levels.
levels(friends_factor) <- grades
friends_factor
```

Notice that the display changed completely.  All the numbers disappeared.  You
only see the labels for each value now.

Notice also that we had to give the labels for the level *in the same order* as
the previous labels.  Previously we had "-1", "1", "2" etc.  We had to specify
the label for "-1" first ("U"), then the label for "1" ("A") and so on.

As you will see, this ability of factors to give labels to categories can be
very useful for the common case where R has columns of data with numerical or
string codes, that refer to categories.  The factor and its levels makes it
much easier to see what the values mean.

## Labeling the UK census data

Now we are going to work on some codes from the [2011 UK
census](https://www.ons.gov.uk/census/2011census/2011censusdata/censusmicrodata/microdatateachingfile/variablelist).
If you follow the link, you will find that the UK census records data about
people as numbers, where the numbers are codes.  For example, it has codes for
marital status, where the codes mean:

* 1=Single (never married or never registered a same-sex civil partnership)
* 2=Married or in a registered same-sex civil partnership 
* 3=Separated but still legally married or separated but still legally in
  a same-sex civil partnership
* 4=Divorced or formerly in a same-sex civil partnership which is now legally
  dissolved
* 5=Widowed or surviving partner from a same-sex civil partnership

Here is a tiny snippet from the full [UK microdata teaching
file](https://www.ons.gov.uk/census/2011census/2011censusdata/censusmicrodata/microdatateachingfile).
The snippet only has 10 rows and 3 columns; the full dataset has many more
rows and columns, as you will soon see.

```{r}
# Load snippet
census_sample <- read.csv('census_snippet.csv')
census_sample
```

Use the `str` function on this data frame to show the column names and data
types:

```{r}
#- Use the "str" function to show column names, types
...
```

In fact these numbers are codes for categories.  We can make this clear to R
by converting the columns to factors.

This is how I convert the `Sex` column to a factor:

```{r}
# Run this chunk
census_sample$Sex <- factor(census_sample$Sex)
census_sample
```

I see from the [variable
list](https://www.ons.gov.uk/census/2011census/2011censusdata/censusmicrodata/microdatateachingfile/variablelist)
that the values in the `Sex` column have these meanings:

* 1=Male
* 2=Female

First I check the current levels:

```{r}
# Run this chunk.
levels(census_sample$Sex)
```

Here is how I set the levels of the `Sex` factor, applying the names in the
same order as the current levels:

```{r}
# Run this chunk.
levels(census_sample$Sex) <- c('Male', 'Female')
census_sample
```

Now your turn.  Make the `Marital.Status` column into a factor:

```{r}
#- Convert Marital.Status column of "census_sample" into a factor.
census_sample$Marital.Status <- ...
# Show the data frame
census_sample
```

Show the current levels of the `Marital.Status` column:

```{r}
#- Show the levels of the Marital.Status column
levels(...)
```

We want new labels for this factor.

* "1" should get the new label "Single"
* "2" should get the new label "Married"
* "3" should get the new label "Separated"
* "4" should get the new label "Divorced"

Make a vector `new_labels` that contains the strings with the new labels above:

```{r}
new_labels <- c(...)
#- Show the vector
new_labels
```

Use `new_labels` to set the labels for the levels of the `Marital.Status`
factor.

```{r}
#- Set the labels of the levels of the 'Marital.Status' factor
levels(...) <- ...
#- Show the data frame
census_sample
```

Now you know factors, you are ready for the full dataset.

## The 2011 census microdata dataset

We are looking at the full version of the sample dataset called the [microdata
teaching
file](https://www.ons.gov.uk/census/2011census/2011censusdata/censusmicrodata/microdatateachingfile)
(MTF).

We will look at the relationships of ethnicity, marital status, estimated
social class and a rating of general health.

Here is the [user guide for the
MTF](https://www.ons.gov.uk/census/2011census/2011censusdata/censusmicrodata/microdatateachingfile/microdatauserguide).

The [MTF variable
list](https://www.ons.gov.uk/census/2011census/2011censusdata/censusmicrodata/microdatateachingfile/variablelist)
gives a list of all the variable in the MTF, with the meaning of their codes.

First we load a complete, unmodified copy of the MTF into R as a data frame.

Be patient - this can take several seconds, because the data file is reasonably
large.

```{r}
# Read the UK 2011 census teaching file.
# The skip=1 option skips the first line of the file, which is
# an information line pointing to the website.
census <- read.csv('2011 Census Microdata Teaching File.csv', skip=1)
# Show the first six rows.
head(census)
```

Show the structure of the data frame with the `str` function:

```{r}
#- Show structure of "census" data frame.
...
```

First we want to drop some rows.

As you will see in the MTF user guide (above), many of the columns in the MTF
file have undefined values for everyone other than "usual residents".

"Usual residents" have a value of 1 in the `Population.Base` column.  We will
start by dropping all the rows but those for the "usual residents".

First we get the `Population.Base` values into a variable for convenience.

```{r}
#- Make a new variable "pop_base" that contains the values
#- of the "Population.Base" column of the "census" data
#- frame.
pop_base <- ...
# Show the first six values of "pop_base"
head(pop_base)
```

Next we make a Boolean vector that we will soon use to select the rows that we
want.

Make a Boolean vector that has `TRUE` in positions where `pop_base` is equal to
1, and `FALSE` otherwise.  Store it in the variable `is_usual`.

```{r}
#- Make a Boolean vector with TRUE in positions where
#- "pop_base" is equal to 1. Store in "is_usual".
is_usual <- ...
# Show the first six values
head(is_usual)
```

Use `subset` to make a new data frame from `census` that only includes the rows
where the corresponding value of `is_usual` is `TRUE`.  Call the new data frame
`usual`.

```{r}
#- Use "subset" to make new data frame called "usual"
#- that is a subset of "census".  "usual" will include
#- only the rows corresponding to "TRUE" values in "is_usual"
usual <- ...
# Show the first six rows of "usual"
head(usual)
```

Among other columns, we are interested in the `Approximated.Social.Grade`.
These are estimated [NRS social
grades](https://en.wikipedia.org/wiki/NRS_social_grade) for each person (and
therefore, for each row).

Here is the explanation of the values in that variable (column) from the MTF
variable list:

* 1=AB
* 2=C1
* 3=C2
* 4=DE
* -9: No code required (People aged under 16, people resident in communal
  establishments, and students or schoolchildren living away during term-time).

First we look at the unique values we have for `Approximated.Social.Grade` in
the `usual` data frame.

Do do this, we use the `table` function.  You'll see more on this in the second Datacamp tutorial.

```{r}
# Use "table" to show the values and counts for the
# "Approximated.Social.Grade" column in "usual"
table(usual$Approximated.Social.Grade)
```

We see that many of the cases in `usual` have -9 as the code for their
`Approximated.Social.Grade`.  One explanation for this, as you see above, is
that the respondent is less than 16.

We can use `table` with more than one vector, to get a cross-tabulation of
counts.  Here we use the `Approximated.Social.Grade` column of `usual`, and
the `Age` category column, to show a cross-tabulation of social grade by age
category.

```{r}
# Use table to show a cross-tabulation of the
# "Approximated.Social.Grade" and "Age" columns of "usual"
table(usual$Approximated.Social.Grade, usual$Age)
```

This should confirm that most of the -9 values in `Approximated.Social.Grade`
are for respondents with value 1 in `Age`.  1 means 0-15 years old; see the
variables list page.

We want to analyze `Approximated.Social.Grade`, so we start by dropping all the
rows with -9 in that column.

The first step is to make a Boolean vector that has TRUE where
`Approximated.Social.Grade` column of `usual` has values greater than 0, and
FALSE otherwise:

```{r}
is_graded <- ...
# Show the first six values
head(is_graded)
```

Use `subset` to put the results into a new data frame, called `ug`, short for
Usual Graded.

```{r}
ug <- subset(...)
# Show the first 6 rows
head(ug)
```

Show a table for `Approximated.Social.Grade` from the `ug` data frame, to
confirm we have lost all the -9 values:

```{r}
table(...)
```

Now we have a somewhat cleaned dataset, we need to make the tables and other
analyses easier to understand, by converting columns to factors.

Convert the following columns of `ug` to factors, using the skills you
acquired earlier in this notebook:

* `Sex`
* `Ethnic.Group`
* `Approximated.Social.Grade`
* `Marital.Status`

The `str` function on the `ug` data frame will show which columns are factors,
when you have done your conversions.

```{r}
#- Convert columns above to factors
ug$Sex <- ...
...
...
...
# Show column types for "ug"
str(ug)
```

Next we want to give readable labels to the levels in our new factors.

For example, here I'm following the steps further up this notebook, to set the
labels (levels) for the `Sex` column.

```{r}
# Run this chunk.
# Set levels for "Sex" column.
levels(ug$Sex) <- c('Male', 'Female')
# Show the levels in table output.
table(ug$Sex)
```

Follow the same recipe to set:

* levels of `Ethnic.Group` to be 'White', 'Mixed', 'Asian', 'Black', 'Chinese/other';
* levels of `Approximated.Social.Grade` to be 'AB', 'C1', 'C2', 'DE';
* levels of `Marital.Status` to be 'Single', 'Married', 'Separated', 'Divorced', 'Widowed'

```{r}
#- Set levels for columns of "ug", as above.
levels(...) <- ...
...
...
# Show the structure of "ug"
str(ug)
```

## The analysis proper

Do social grades differ between ethnic groups?

Use `table` to make a cross-tabulation of counts by `Ethnic.Group` and
`Approximated.Social.Grade`.  Store this table as `eg_by_sg`.

```{r}
eg_by_sg <- ...
# Display the table.
eg_by_sg
```

The `mosaicplot` function is a very nice way of showing proportions in
cross-tabulations.  It shows each entry (cell) in the table as a rectangle.  The area of the rectangle shows the proportion of the values for the whole column that are contained in this cell.

This is easiest to see by example.  Here we use `mosaicplot` to do a mosaic
plot of `eg_by_sg`.  You will see more on these in the second Datacamp
tutorial.

```{r}
mosaicplot(eg_by_sg)
```

For reflection - what conclusions can you draw about ethnic group and social
class?

Does social grade differ by marital status?

```{r}
#- Make a new table "ms_by_sg" that is the cross-tabulation of
#- marital status by social grade.
ms_by_sg <- ...
# Show "ms_by_sg"
ms_by_sg
```

Do a mosaic plot of `ms_by_sg`

```{r}
#- Mosaicplot of "ms_by_sg"
...
```

For reflection - what conclusions can you draw about the effect of marital
status on social grade, if any?

## Does health differ by social grade?

Here are the codes for the `Health` column of the dataset:

* 1=Very good health
* 2=Good health
* 3=Fair health
* 4=Bad health
* 5=Very bad health
* -9: not coded.

```{r}
#- Make a table to cross-tabulate `Health` against social grade.
h_by_sg <- ...
# Display table with a mosaicplot
mosaicplot(h_by_sg)
```

For reflection - what conclusions can you draw?

What is the relationship of health and ethnic group?  Do a mosaic plot to investigate:

```{r}
#- mosaicplot of health against social group
...
```

Make a mosaic plot to show the relationship of health and gender (encoded in
the `Sex` column of the data frame).

```{r}
#- mosaicplot of health against sex
...
```

## In the West Midlands

If you have time left over, use `subset` to select only the respondents from
the West Midlands.  See the variable list page at the top for the values you
will need to do this.  Call this data frame `ug_wm`.

Look at the relationship of ethnic group and social grade for this dataset.  Does it differ from the pattern of the whole population?

How about the relationship of health and social group?